Generating Correlated Data for Omics Simulation

### 2025-08-05 manuscript/paper.qmd
Second review revisions

Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real, reference dataset and producing simulations that mimic it.
However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
For Guassian copulas, this is a choice of a covariance matrix.
However, the sample covariance matrix of a reference dataset is a poor choice and we present three alternative choices.

@@

```{r}
#| label: fig-intro
#| fig-cap: "(a) Schematic of our Gaussian copula-based approach. The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form. (b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} := \\widehat \\Sigma_{\\text{ref}}$. 100 genes and 12 samples were used for both reference and the 20 simulated datasets. Both reference and simulated data follow a multivariate normal distribution. Eigenvalues are numbered from largest to smallest. (c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, coprcor), SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
#| fig-width: 8
#| fig-height: 6

@@

# Plot benchmarking
METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula", "mvrnorm")
METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
benchmark <- read_tsv("../processed/benchmark/results.txt") |>
    filter(method != "indep") |> # We don't want to compare indep for this, whose runtime hasn't been optimized anyway
    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints

@@

We note that, unlike many timeseries methods that consider correlation across time points, the dependence here is between genes within the same sample and no dependence between samples or timepoints is generated.
This reflects scenarios where CYCLOPS is used, for example, post-mortem biopsy samples of unrelated individuals collected at different times [@CYCLOPS2].

@@

Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
We use two basic facts about the multivariate distribution.
First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma_{\text{sim}}) $$
as desired.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used (more fine-grained buckets will produce higher correlation).
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

Other Gaussian copula-based R packages include `copula`, `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
Metabolomics data was obtained from MetaboAnalyst (https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv).

>>>>

Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real, reference dataset and producing simulations that mimic it.
However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
For Gaussian copulas, this is a choice of a covariance matrix.
However, the sample covariance matrix of a reference dataset is a poor choice and we present three alternative choices.

@@

```{r}
#| label: fig-intro
#| fig-cap: "(a) Schematic of our Gaussian copula-based approach. The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form. (b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} := \\widehat \\Sigma_{\\text{ref}}$. 100 genes and 12 samples were used for both reference and the 20 simulated datasets. Both reference and simulated data follow a multivariate normal distribution. Eigenvalues are numbered from largest to smallest. (c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, corpcor), SPsimSeq, mvrnorm (from the MASS package), the mvnfast package, and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts. All methods used a single thread."
#| fig-width: 8
#| fig-height: 6

@@

# Plot benchmarking
METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula", "mvrnorm", "mvnfast")
METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
benchmark <- read_tsv("../processed/benchmark/results.txt") |>
    filter(method != "indep") |> # We don't want to compare indep for this, whose runtime hasn't been optimized anyway
    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints

@@

We note that, unlike many timeseries methods that consider correlation across time points, the dependence here is between genes within the same sample and generated samples are independent, conditional upon the time point.
This reflects scenarios where CYCLOPS is used, for example, post-mortem biopsy samples of unrelated individuals collected at different times [@CYCLOPS2].

@@

Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
We use two basic facts about the multivariate normal distribution.
First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$ are independent, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
Therefore, given matrices $D, U, W$ and independent random vectors $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma_{\text{sim}}) $$
as desired.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020], due to its popularity, support for bulk RNA-Seq, and use of Gaussian copulas.
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used (more fine-grained buckets will produce higher correlation).
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

Other Gaussian copula-based R packages include `copula`, `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases due to the reliance on the Cholesky decomposition, see @fig-intro d.

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
Metabolomics data was obtained from MetaboAnalyst ([https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv](https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv)).
